A note on observation processes in epidemic models

### 2020-02-13 ms/response.tex

\revtext{Authors aim to evaluate how using the wrong model in terms of the distributions of the infectious and latent periods affect estimates of $\mathcal R_0$.}

\revtext{Authors did some work in this direction, but I believe more work is needed to address their question. In particular, authors will need to use a model that incorporates realistic distributions of latent and infectious periods. They can accomplish this using the linear chain tricks or alternative integro-differential equation models. Then they can generate simulated data from that "true/realistic" model and go ahead and estimate $\mathcal R_0$ using simple models as they do here.}

\revtext{Having said this, I would like to point out that I am sure there are a number of papers out there that have look at similar questions in the context of different infectious diseases, etc. It'd be good to make sure the relevant literature is considered and appropriately cited.}

The reviewer may have misconstrued our goals in this paper. We were \emph{not} exploring the effects of the process delay distributions (i.e. latent and infectious periods), and their mis-specification, on epidemiological dynamics and parameter estimation. Rather, we were exploring the effects of reporting delays (and their mis-specification) on parameter estimation.

However, reporting delays do depend implicitly on the process delays. We have added the following paragraph in the Discussion to address this point:

@@

``Delay distributions in reporting time implicitly depend on latent and infectious 
period distributions. For example, when cases are assumed to be reported
at the end of infectious periods,
the delay distribution is equivalent to the convolution of the latent and
infectious period distributions. These distributions play important roles in 
shaping the epidemiological dynamics, including the probability and size of an outbreak, 
the stability and persistence of a system, and the 
prediction of dynamical transitions. Assumptions about
latent and infectious period distributions can also have large effects on the
estimates of the basic reproductive number.''

Although we could extend the paper to consider dynamics (and reporting time delays) based on more realistic distributions, our goal here was to evaluate the effects of these assumptions qualitatively, in the simplest possible epidemiological model.

@@

\revtext{I have one comment that the authors may wish to consider, although this doesn't rise to the level of a critical flaw that must be corrected.  It seemed to me that the Discussion lost sight of the fact that statistical models often have to make simplifying assumptions.  Despite these simulation results, it seems to me that making a simplifying assumption about the process generating the incidence data may not be a terrible thing to do with real data.  When it comes to fitting statistical models to noisy data, models that are "more realistic", in the sense of incorporating more of the underlying mechanism, do not necessarily lead to improved inference.  One has to choose their battles with respect to what one tries to capture in a model, and what one simplifies away.  With real data, it is entirely unclear whether model fit would be improved by fitting a model with a more complicated description of the process generating the incidence data, and it may very well be that such a
question can only be answered on a case-by-case basis.  In the absence of such understanding, it seems to me that the Discussion could strike a somewhat more measured tone regarding the merits of simpler vs. more complicated models.}

We agree. We did not intend to say that modelers should always strive to build the more complex models. We have revised the final paragraph of the Discussion to better reflect our intentions:

### 2020-02-13 ms/ms.tex
stuff
i0, small wording

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/example.pdf}
\caption{
\textbf{A comparison of incidence measured at two different time points in infection.}
(A) 
A deterministic simulation of the SIR model using the following parameters: 
$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
and $\Delta t = 0.1$ time units.
(B-C) Effects of $\mathcal R_0$ on the relative difference in the size of the epidemic peak ($1 - [\max i_2(t)]/[\max i_1(t)]$) and the difference in the peak timing ($\hat{t}_2 - \hat{t}_1$ where $i_k(\hat{t}_k) = \max i_k(t)$ for $k = 1, 2$) due to delays in reporting time.
Open circles: simulation results of the deterministic SIR model.
Solid lines: locally estimated scatterplot smoothing (LOESS) curves fitted to simulation results.
Remaining parameters are held constant ($1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$, and $\Delta t = 0.1$ time units) throughout simulations.
}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{../figure/compare_deterministic.pdf}
\caption{
\textbf{A comparison of incidence measured at two different time points in infection.}
We simulate infection time series and recovery time series 100 times using the 
following parameters:  
$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
$\rho = 0.5$, $\theta = 10$, and $\Delta t = 0.1$ time units.
For each simulation, we fit the infection model and the recovery model by
(A) estimating the mean infectious period and (B) assuming
that the mean infectious period is known.
Coverage probabilities represent the proportion of confidence intervals
that contain the true value of the basic reproductive number ($\mathcal R_0$).
}
\end{figure}

@@

@@

Our study shows that a seemingly negligible change in the assumptions of an epidemic
model can affect the inference of infectious disease transmission.
While the amount of bias introduced from the misspecification of the observation
model can be fairly small (e.g., less than 5\% in our examples), any modelers
trying to make serious forecasts should be aware of potential biases.
We caution disease modelers to carefully consider the implications of their model assumptions when developing epidemic models.

>>>>

\begin{figure}[!t]
\includegraphics[width=\textwidth]{example.pdf}
\caption{
\textbf{A comparison of incidence measured at two different time points in infection.}
(A) 
A deterministic simulation of the SIR model using the following parameters: 
$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i_0 = 1 \times 10^{-4}$,
and $\Delta t = 0.1$ time units.
(B-C) Effects of $\mathcal R_0$ on the relative difference in the size of the epidemic peak ($1 - [\max i_2(t)]/[\max i_1(t)]$) and the difference in the peak timing ($\hat{t}_2 - \hat{t}_1$ where $i_k(\hat{t}_k) = \max i_k(t)$ for $k = 1, 2$) due to delays in reporting time.
Open circles: simulation results of the deterministic SIR model.
Solid lines: locally estimated scatterplot smoothing (LOESS) curves fitted to simulation results.
Remaining parameters are held constant ($1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i_0 = 1 \times 10^{-4}$, and $\Delta t = 0.1$ time units) throughout simulations.
}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{compare_deterministic.pdf}
\caption{
\textbf{A comparison of incidence measured at two different time points in infection.}
We simulate infection time series and recovery time series 100 times using the 
following parameters:  
$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i_0 = 1 \times 10^{-4}$,
$\rho = 0.5$, $\theta = 10$, and $\Delta t = 0.1$ time units.
For each simulation, we fit the infection model and the recovery model by
(A) estimating the mean infectious period and (B) assuming
that the mean infectious period is known.
Coverage probabilities represent the proportion of confidence intervals
that contain the true value of the basic reproductive number ($\mathcal R_0$).
}
\end{figure}

@@

Distributions of reporting time delays implicitly depend on the latent and infectious 
period distributions. For example, when cases are reported
at the end of infectious periods \citep{breto2009time, he2009plug, lin2016seasonality},
the delay distribution is equivalent to the convolution of the latent and
infectious period distributions. These distributions also play important roles in 
shaping the epidemiological dynamics,
including the probability and size of an outbreak
\citep{anderson1980spread}, the stability and persistence of a system
\citep{keeling1998effect, lloyd2001destabilization, lloyd2001realistic},
and the  occurrence of dynamical transitions \citep{krylova2013effects}.
Assumptions about latent and infectious period distributions can also have large effects on the estimates of the basic reproductive number \citep{wearing2005appropriate}.

@@

All statistical models rely on simplifying assumptions.
Given a range of possible assumptions, modelers have to
decide which aspects of a system to capture 
at the cost of simplifying other aspects.
However, some assumptions are often taken for granted.
Epidemic modelers tend to focus on assumptions about the
transmission process; processes through which cases are reported are often 
overlooked and simplified.
Our study shows that a seemingly negligible change in the assumptions of an epidemic
model can affect the inference of 
infectious disease transmission.
Although the amount of bias introduced from the misspecification of the observation
model can be small (e.g., less than 5\% in our examples), modelers
trying to make serious predictions may want to control for potential biases.
We caution disease modelers to be more careful about their modeling decisions as they may have unexpected consequences on the inference.
